### 2023-02-07 paper-thesis/chapters/03-background-dependent-types.tex
Start reworking spoofax chapter
Improve chapter 3

This chapter will explain the concepts behind dependent types. If you are already familiar with Dependent Types, you can skip this chapter.

\section{What are dependent types?}

@@

The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example it is now possible to express a function that adds two vectors of the same length and returns a new vector of that length. 
\begin{lstlisting}
add_vec : (n : Nat) -> Vec n -> Vec n -> Vec n
\end{lstlisting}

\section{The Curry-Howard Isomorphism}

The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs~\cite{chc}. This is done by representing false statements as empty types, and true statements as non-empty (inhabited) types.

For example, it is possible to define a type \verb|IsEven n|, that represent the statement "\verb|n| is even", and declare it in such a way that it is only inhabited if \verb|n| is even.

In order to use a type system to prove things it is important that the type system is sound, that is, it is impossible to prove a false statement. Using the Curry-Howard correspondence, that means it should be impossible to obtain an element of the empty type. We will explore this further in chapter \ref{ch:universes}.

\section{The Lambda Cube}

The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other~\cite{lambda_cube}. Specifically, it categorizes languages on three axis: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}

@@

The \textit{Calculus of Constructions} is the language that has all three of these features, and it is what we will implement in this paper. If you want to learn more, I recommend playing around with the \textit{\href{https://agda.readthedocs.io/en/latest/}{Agda}} language, which is a real-world language that satisfies all three axis.

>>>>

This chapter will explain the concepts behind dependent types. Readers already familiar with Dependent Types are recommended to only skim through this chapter.

\section{What are Dependent Types?}

@@

The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example:
\begin{lstlisting}
append : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n -> Vec A (n + n)
\end{lstlisting}
A \verb|Vec A n| is a list with elements of type \verb|A| that is exactly length \verb|n| (an integer, which is a value!). This append function appends two \verb|Vecs| of equal length, returning a \verb|Vec| of length \verb|n + n|. This is an example of a type-level computation.

Dependent types are useful because dependent type systems allow for the Curry-Howard Correspondence~\cite{chc}. Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition. Thus if our type system is sound\footnote{The type system we will define in chapter \ref{chap:baselang} is not sound, we will fix this in chapter \ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.

\section{The Calculus of Constructions}

The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other~\cite{lambda_cube}. Specifically, it categorizes languages on three axes: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}

@@

The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter \ref{chap:baselang}.
